Probing the inner kpc of massive galaxies with strong gravitational lensing

### 2015-01-06 response.txt

We are grateful to the referee for careful reading of the 
manuscript and helpful suggestions.

@@

==================================================================================

What was the source size used to generate Figure 1?

>> RESPONSE: The source has a Gaussian profile with FWHM of 1170 pc 
>> (rms of 500 pc). We have added this information to the caption:

" a background source with a Gaussian light profile (FWHM 1170 pc)..."

@@

==================================================================================

How was the black hole mass chosen? I assume it follows the M-sigma 
relation, but that is not discussed explicitly (perhaps because the 
effective velocity dispersion of the model galaxy is not 
mentioned).

>> RESPONSE: For a lens galaxy with a sigma of 200 km/s the black hole
mass on the M-sigma relation is about 2e8 Msun (from Kormendy & Ho 2013
and McConnell et al. 2011). The giant elliptical strong lenses have typical
velocity dispersions of  200 km/s (Sonnenfeld et al. 2013), so we think that 
black holes in the 10^8 Msun mass range are plausible for these galaxies.
We have added this information to the simulation section:

"Lensing galaxies are typically massive ellipticals with velocity dispersions 
of order 200 km/s (e.g., Sonnenfeld et al. 2013). From the M-σ relation, we predict
that these galaxies should contain black holes with masses in excess 
of 2 x 10^8Msun (e.g., Kormendy & Ho 2013). In our simulations we have chosen 
black holes with masses of 1, 2, and 4 x 108 Msun ."
==================================================================================

@@

"The stellar population is modeled as an elliptical cored 
power-law. The deflection angles for elliptical cored power-law 
models were calculated using FastEll (Barkana 1998)."

==================================================================================

"For double and quad image configurations, the theory of strong 
lensing predicts a third and a fifth image near the centers of 
lensing galaxies." Strictly speaking, that statement applies only 
under certain conditions. Even if those conditions are thought to 
apply to realistic galaxy models, it is important to understand 
that the theoretical statement is nuanced. For example, there is no 
central image if the surface mass density has a central cusp that 
is steeper than 1/r, or if the SMBH exceeds a threshold in mass. If 
the black hole mass is below that threshold, there can actually be 
more than one central image.

>> RESPONSE: We have tried to qualify this with the sentences that
follow this description:

"For double and quad image configurations, the theory of strong lensing 
predicts a third and a fifth image near the centers of lensing galaxies. 
However, realistic galaxy density profiles produce central images that are 
almost always demagnified, making their detection difficult (e.g., Jackson 2013). 
If the lens galaxies have cuspy central profiles with steep slopes or very 
massive SMBHs, this central image may not form at all."

### 2015-01-06 ms.tex
final revisions for resubmission

Strong gravitational lensing is a powerful tool for probing the matter distribution in distant galaxies. Among other things, strong lenses have been used to constrain galaxy masses, density profiles, and abundance of dark matter subhalos \citep[e.g.,][]{dalal:02, Gavazzi:07, Bolton:08}. For double and quad image configurations, the theory of strong lensing predicts a third and a fifth image near the centers of lensing galaxies. 
\blue{Realistic galaxy density profiles produce central images that are almost always \emph{demagnified}\footnote{approximately as $(\Sigma/\Sigma_{crit})^{-2}$, where $\Sigma$ is the surface density and $\Sigma_{crit}$ is the critical surface density}, making their detection difficult \citep[e.g.,][]{Jackson:13}. If the lens galaxies have cuspy central profiles with steep slopes or very massive SMBHs, this central image may not form at all.}  
%\red{Unlike the other lensed images, which are magnified, this image can be significantly \emph{demagnified}, making its detection difficult \citep[e.g.,][]{Jackson:13}.}
 It is well-understood that the magnification of the central images is very sensitive to the matter distribution in the innermost regions of lens galaxies: very steep density profiles significantly demagnify the central images, whereas cored or shallow profiles render them brighter (Figure \ref{fig1}).

@@

We modeled the lens potential as a sum of three components: dark matter, stellar population and a single central SMBH. The dark matter component was modeled as \blue{an elliptical} singular power-law with a slope of 0.1, in agreement with the projected surface mass density of the NFW profile in the innermost regions \citep{Golse:02}. As  pointed out by \citet{Keeton:03}, due to its expected flatness in these regimes, the dark matter component has negligible influence on the central images (although models with dark matter cusps have been entertained). 
The stellar population is modeled as \blue{an elliptical} cored power-law, $\Sigma(r)\propto (r^2+R_{\mathrm{core}}^2)^{\gamma_s}$. \blue{The deflection angles for elliptical cored power-law models were calculated using FastEll \citep{Barkana:98}.}  Figure \ref{fig2} shows the stellar component (red curve), dark matter component (black dashed curve) and the sum of the two (black solid curve). The grey dashed curves show a few examples of core-sersic models with parameters taken from \citet{Ferrarese:06}. Although the cored power-law model used in this work does not account for the slope of the stellar distribution below the core break, it is a close fit and a reasonable model to approximate the observed stellar light profiles in the inner few kpc. 
The black hole was modeled as a simple point mass at the center of the potential. We assumed that all the three components are concentric, though we note that two- and three-black hole interactions may invalidate this assumption.
The stellar and dark mass of the lens were normalized such that they each contain $1.5\times10^{11} M_{\odot}$ in a radius of 10 kpc. The resulting Einstein radius is of order 1 arcsec, in agreement with galaxy-galaxy strong lenses \citep[e.g.,][]{hezaveh:13b}.
To predict the visibilities, we calculated the ALMA $uv$-coverage of a 10-$hr$ long observation (full array), using the $simobserve$ task of Common Astronomy Software Applications package, for an observing frequency of 150 GHz. 
The model visibilities were calculated by computing the 2D Fourier transform of the surface brightness data maps and resampling the Fourier transform maps over the $uv$-coverage. 
The noise was estimated using ALMA sensitivity calculator for a channel width of 400 km/s at 150 GHz.

@@

\begin{figure}
\begin{center}
\centering
\includegraphics[trim= 0 0 0 0, clip,width=0.48\textwidth]{figures/f_03.eps}
\centering
\end{center}
\caption{ Detection significance of a central image, as a function of the stellar core size for a 10-hr long ALMA observation. 
The fiducial model (solid black curve) corresponds to $\gamma_{stellar}=1.0$, $M_{\mathrm{BH}}=2\times10^8 M_{\odot}$, and a source size of 500 pc (radius). The dashed black curve shows the detection significance when the SMBH is removed and the solid red curve shows the effect of a steeper density profile ($\gamma_{\mathrm{stellar}}=1.1$). The solid blue and gray dotted curves show the detections for a source with a radius of 1 kpc and 250 pc respectively. 
\label{fig:3}}
\end{figure}

>>>>

Strong gravitational lensing is a powerful tool for probing the matter distribution in distant galaxies. Among other things, strong lenses have been used to constrain galaxy masses, density profiles, and abundance of dark matter subhalos \citep[e.g.,][]{dalal:02, Gavazzi:07, Bolton:08}. For double and quad image configurations, the theory of strong lensing predicts a third and a fifth image near the centers of lensing galaxies. 
\blue{However, realistic galaxy density profiles produce central images that are almost always \emph{demagnified}\footnote{approximately as $(\Sigma/\Sigma_{crit})^{-2}$, where $\Sigma$ is the surface density and $\Sigma_{crit}$ is the critical surface density}, making their detection difficult \citep[e.g.,][]{Jackson:13}. If the lens galaxies have cuspy central profiles with steep slopes or very massive SMBHs, this central image may not form at all.}  
%\red{Unlike the other lensed images, which are magnified, this image can be significantly \emph{demagnified}, making its detection difficult \citep[e.g.,][]{Jackson:13}.}
 It is well-understood that the magnification of the central images is very sensitive to the matter distribution in the innermost regions of lens galaxies: very steep density profiles significantly demagnify the central images, whereas cored or shallow profiles render them brighter (Figure \ref{fig1}).

@@

We modeled the lens potential as a sum of three components: dark matter, stellar population and a single central SMBH. The dark matter component was modeled as \blue{an elliptical} singular power-law with a slope of 0.1, in agreement with the projected surface mass density of the NFW profile in the innermost regions \citep{Golse:02}. As  pointed out by \citet{Keeton:03}, due to its expected flatness in these regimes, the dark matter component has negligible influence on the central images (although models with dark matter cusps have been entertained). 
The stellar population is modeled as \blue{an elliptical} cored power-law, $\Sigma(r)\propto (r^2+R_{\mathrm{core}}^2)^{\gamma_s}$. \blue{The deflection angles for elliptical cored power-law models were calculated using FastEll \citep{Barkana:98}.}  Figure \ref{fig2} shows the stellar component (red curve), dark matter component (black dashed curve) and the sum of the two (black solid curve). The grey dashed curves show a few examples of core-sersic models with parameters taken from \citet{Ferrarese:06}. Although the cored power-law model used in this work does not account for the slope of the stellar distribution below the core break, it is a close fit and a reasonable model to approximate the observed stellar light profiles in the inner few kpc. 
The black hole was modeled as a simple point mass at the center of the potential. 
\blue{Lensing galaxies are typically massive ellipticals with velocity dispersions of order 200 km/s \citep[e.g.,][]{Sonnenfeld:13}. From the M-$\sigma$ relation, we predict that these galaxies should contain black holes with masses in excess of $2\times10^8 M_{\odot}$ \citep[e.g.,][]{Kormendy:13}. In our simulations we have chosen black holes with masses of 1, 2, and $4 \times 10^8 M_{\odot}$.}

We assumed that all the three components are concentric, though we note that two- and three-black hole interactions may invalidate this assumption.
The stellar and dark mass of the lens were normalized such that they each contain $1.5\times10^{11} M_{\odot}$ in a radius of 10 kpc. The resulting Einstein radius is of order 1 arcsec, in agreement with galaxy-galaxy strong lenses \citep[e.g.,][]{hezaveh:13b}.
To predict the visibilities, we calculated the ALMA $uv$-coverage of a 10-$hr$ long observation (full array), using the $simobserve$ task of Common Astronomy Software Applications package, for an observing frequency of 150 GHz. 
The model visibilities were calculated by computing the 2D Fourier transform of the surface brightness data maps and resampling the Fourier transform maps over the $uv$-coverage. 
The noise was estimated using ALMA sensitivity calculator for a channel width of 400 km/s at 150 GHz.

@@

\begin{figure}
\begin{center}
\centering
\includegraphics[trim= 0 0 0 0, clip,width=0.48\textwidth]{figures/f_03.eps}
\centering
\end{center}
\caption{ Detection significance of a central image, as a function of the stellar core size for a 10-hr long ALMA observation. 
The fiducial model (solid black curve) corresponds to $\gamma_{stellar}=1.0$, $M_{\mathrm{BH}}=2\times10^8 M_{\odot}$, and a source size of 500 pc (radius). \blue{The lens has an ellipticity of 0.2.} The dashed black curve shows the detection significance when the SMBH is removed and the solid red curve shows the effect of a steeper density profile ($\gamma_{\mathrm{stellar}}=1.1$). The solid blue and gray dotted curves show the detections for a source with a radius of 1 kpc and 250 pc respectively. 
\label{fig:3}}
\end{figure}
